Reasoning About a Machine with Local Capabilities

### 2018-05-04 toplas/paper.tex
Added first draft for section about reasoning

% Conv defs
\newcommand{\lookingat}[3]{\ensuremath{#1} \text{ is looking at } \ensuremath{#2} \text{ followed by } \ensuremath{#3}}
\newcommand{\pointstostack}[3]{\ensuremath{#1} \text{ points to stack with } \ensuremath{#2} \text{ used and } \ensuremath{#3} \text{ unused}}
\newcommand{\nonlocal}[1]{\ensuremath{#1} \text{ is non-local}}

@@

\begin{figure}[htb]
  \begin{toplas}
    \centering
\begin{lstlisting}
// push private registers to the stack
  push $r_{\var{priv},1}$
  ...
  push $r_{\var{priv},n}$
// push restoration code to the stack
  push $\encode(i_1)$
  push $\encode(i_2)$
  push $\encode(i_3)$
  push $\encode(i_4)$
// push old pc to the stack
  move r_t1 pc
  lea r_t1 $\var{ofc}$ 
  push r_t1
// push stack pointer to the stack
  push r_stk
// set up protected return pointer
  move r_0 r_stk
  lea r_0 $\var{ofc_\var{act}}$ // -5 is the offset to the first instruction of the activation code
  restrict r_0 $\encodePermPair((\local,\entry))$
// restrict stack capability to unused part
  geta r_t1 r_stk
  plus r_t1 r_t1 1
  getb r_t2 r_stk
  subseg r_stk r_t1 r_t2
// clear unused part of the stack
  mclear r_stk
// clear non-argument registers
  rclear $R$
  jmp r
after:
// pop the restore code
  pop r_t1
  pop r_t1
  pop r_t1
  pop r_t1
// pop the private state into approriate registers
  pop $r_{\var{priv},1}$
  ...
  pop $r_{\var{priv},n}$
\end{lstlisting}
    \caption{Implementation of \texttt{scall} $r(r_{\var{args},1},\dots,
      r_{\var{args},m},r_{\var{priv},1},\dots, r_{\var{priv},n})$. The
      restoration code is given in \figurename~\ref{fig:scall-rest-code}. The
      variable $\var{ofc}$ is the offset to the label after, and
      $\var{ofc_\var{act}} = -5$ which is the offset to the first instruction of
      the activation record. The set $R = \RegName \setminus
      \{\texttt{pc},\texttt{r\_stk},\texttt{r\_0},r,r_{\var{args},1},\dots,r_{\var{args},m}\}$.}
    \label{fig:scall-impl}
  \end{toplas}

@@

\begin{toplas}
\section{Reasoning about programs}
\label{sec:reasoning}
\itoplassug{
  \begin{itemize}
  \item A section with the lemmas we use for reasoning about programs
    \begin{itemize}
    \item lemma 58, scall works. Include description of how it handles the
      common things in call and leaves the argumentation for callee code and
      continuation.
    \item lemma 59, malloc works. If this is included, we should probably also
      include the malloc specification.
    \item lemma 60, create closure works.
    \item Include a description about how the FTLR is used to reason about
      unknown code.
    \item Mention that this is a step on the way towards having a program logic,
      and then in the discussion discuss why a program logic for this language
      is not what we need.
    \end{itemize}
  \end{itemize}
}

\end{toplas}

@@

>>>>

% Conv defs
\newcommand{\linksto}[4]{\ensuremath{#1} \text{ links } \ensuremath{#2} \text{ as } \ensuremath{#3} \text{ to } \ensuremath{#4}}
\newcommand{\lookingat}[3]{\ensuremath{#1} \text{ is looking at } \ensuremath{#2} \text{ followed by } \ensuremath{#3}}
\newcommand{\pointstostack}[3]{\ensuremath{#1} \text{ points to stack with } \ensuremath{#2} \text{ used and } \ensuremath{#3} \text{ unused}}
\newcommand{\nonlocal}[1]{\ensuremath{#1} \text{ is non-local}}

@@

\begin{figure}[htb]
  \begin{toplas}
    \centering
\begin{lstlisting}
// push private registers to the stack
  push $r_{\var{priv},1}$
  ...
  push $r_{\var{priv},n}$
// push restoration code to the stack
  push $\encode(i_1)$
  push $\encode(i_2)$
  push $\encode(i_3)$
  push $\encode(i_4)$
// push old pc to the stack
  move r_t1 pc
  lea r_t1 $\var{ofc}$ 
  push r_t1
// push stack pointer to the stack
  push r_stk
// set up protected return pointer
  move r_0 r_stk
  lea r_0 $\var{ofc_\var{rec}}$ // -5 is the offset to the first instruction of
the recovery code
  restrict r_0 $\encodePermPair((\local,\entry))$
// restrict stack capability to unused part
  geta r_t1 r_stk
  plus r_t1 r_t1 1
  getb r_t2 r_stk
  subseg r_stk r_t1 r_t2
// clear unused part of the stack
  mclear r_stk
// clear non-argument registers
  rclear $R$
  jmp r
after:
// pop the restore code
  pop r_t1
  pop r_t1
  pop r_t1
  pop r_t1
// pop the private state into approriate registers
  pop $r_{\var{priv},1}$
  ...
  pop $r_{\var{priv},n}$
\end{lstlisting}
    \caption{Implementation of \texttt{scall} $r(r_{\var{args},1},\dots,
      r_{\var{args},m},r_{\var{priv},1},\dots, r_{\var{priv},n})$. The
      restoration code is given in \figurename~\ref{fig:scall-rest-code}. The
      variable $\var{ofc}$ is the offset to the label after, and
      $\var{ofc_\var{rec}} = -5$ which is the offset to the first instruction of
      the activation record. The set $R = \RegName \setminus
      \{\texttt{pc},\texttt{r\_stk},\texttt{r\_0},r,r_{\var{args},1},\dots,r_{\var{args},m}\}$.}
    \label{fig:scall-impl}
  \end{toplas}

@@

\begin{toplas}
\section{Reasoning about programs on a capability machine}
\label{sec:reasoning}
\itoplassug{
  \begin{itemize}
  \item A section with the lemmas we use for reasoning about programs
    \begin{itemize}
    \item lemma 58, scall works. Include description of how it handles the
      common things in call and leaves the argumentation for callee code and
      continuation.
    \item lemma 59, malloc works. If this is included, we should probably also
      include the malloc specification.
    \item lemma 60, create closure works.
    \item Include a description about how the FTLR is used to reason about
      unknown code.
    \item Mention that this is a step on the way towards having a program logic,
      and then in the discussion discuss why a program logic for this language
      is not what we need.
    \end{itemize}
  \end{itemize}
}

% Anti-reduction lemma
Out capability machine only allows the final memory of an execution to be
observed, so naturally the correctness lemmas we prove in
\sectionname~\ref{sec:examples} are statements about the memory in the halted
configuration. In order to prove something about the final memory, we encode it
as a region in a world and show that the initial configuration is in the
$\observations$-relation w.r.t.\ that world. The $\observations$-relation says
that if the initial configuration halts successfully, then the memory in the
halting state must still respect a private future world of the initial world. It
is, however, bothersome and error prone to try to argue about the entire
execution in one go. Instead we want to reason modularly, in the sense that we
only want to reason about parts of the execution at a time. To this end, we
prove an anti-reduction lemma that essentially says that if we can show that
an initial configuration, $\Phi$, steps to a configuration $\Phi'$ and
that $\Phi'$ is in the $\observations$-relation, then $\Phi$ is also in the
observation relation.
\begin{lemma}[Anti-reduction for $\observations$]
\label{lem:anti-red-obs}
  \begin{align*}
    & \forall n, n',i,\reg, \reg', \ms, \ms', \ms_r, W, W' \ldotp\\
    & \quad n' \geq n - i \land W' \futurestr W \land \\
    & \quad (\forall \ms_f \ldotp (\reg,\ms \uplus \ms_r \uplus \ms_f) \step[i] (\reg',\ms' \uplus \ms_r \uplus \ms_f)) \land \\
    & \quad \npair[n']{(\reg',\ms')} \in \observations(W') \\
    & \qquad \Rightarrow \npair{(\reg,\ms \uplus \ms_r)} \in \observations(W)
  \end{align*}
\end{lemma}

% FTLR
The anti-reduction works well when we reason about the execution of known code,
because we can execute it. When we want to reason about unknown code, the
anti-reduction lemma does not apply because we do not know what instructions are
being executed. This is where the FTLR (Theorem~\ref{thm:ftlr}) comes into play.
As a reminder, the FTLR says that if a capability only has access to safe values
with respect to a world, then it is safe to use it for execution in the same
world. The unknown code we consider will be assumed to only have access to safe
values, which typically means that we assume it has access to a linking table
with a malloc capability and otherwise consists of instructions. This allows us
to use the FTLR seeing as Malloc is a safe value cf.
Lemma~\ref{lem:malloc-in-vr} and instructions are really just integers and they
are always safe values. Here, it is important to remember that the FTLR gives us
that the capability for the untrusted code is safe w.r.t. some world, so even
though we described the usual assumption on the unknown code's memory, it is the
semantic model of the memory, i.e.\ the world, that the FTLR considers. In order
to argue that a specific configuration is safe, we need to argue that the
configuration is safe with respect to the world. That is, we need to argue that
the memory satisfies the world and the register-file is in the
$\stdrr$-relation. For instance, in the case where the untrusted code assumes
control first, we will use the FTLR to show that the capability for the unknown
code can be used for execution. The unknown code will get access to some known,
trusted code through a capability in the initial register-file. Because the
capability is in the register-file, we will have to argue that it is safe which
we do using Lemma~\ref{lem:anti-red-obs}.

% Many details to get right when reasoning about low-level programs.
There are many details to get right when programming in assembly. These details
carry over to proofs about assembly programs, so the proofs become cumbersome.
It is especially annoying, when the same line of reasoning is applied in
multiple places. For instance, in $\texttt{scall}$ the setup before the jump to
the call is always the same and the recovery upon return is also always the
same, and in both cases the reasoning will be the same every time.
We deal with this part of the reasoning in the following lemma.
\begin{lemma}[$\mathtt{scall}$ works]
  \label{lem:scall-works}
  If
  \begin{enumerate}
  \item \label{item:scall:memsat} $\memSat[n]{\ms}{\revokeTemp{W}}$ 
  \item $\dom(\ms_f) \cap (\dom(\ms_\stk \uplus \ms_{\mathit{unused}} \uplus \ms)) = \emptyset$
  \item \label{item:scall:stack} $(\reg,\ms) \text{ is looking at }
    \mathtt{scall}\;r(\overline{r_{\mathit{arg}}},
    \overline{r_{\mathit{priv}}}) \text{ followed by }
    c_{\mathit{next}}$\footnote{Defined in the appendix.}
  \item $\reg \text{ points to stack with $\ms_\stk$ used and $\ms_{\mathit{unused}}$ unused}$
  \item \label{item:scall:hyp-callee} \emph{Hyp-Callee} If
    \begin{itemize}
    \item $\dom(\ms_{\mathit{unused}}) = \dom(\ms_{\mathit{rec}}
      \uplus \ms_{\mathit{unused}}')$,
    \item $W' =
      \revokeTemp{W}[\iota^{\sta}(\temp,\ms_\stk\uplus\ms_{\mathit{rec}} \uplus \ms_f),\iota^{\pwl}(\dom(\ms_{\mathit{unused}}'))]$,
    \item $\memSat[n-1]{\ms''}{W'}$
    \item $\reg' \text{ points to stack with $\emptyset$ used and $\ms_{\mathit{unused}}'$ unused}$
    \item $\reg'= \reg_0[\pcreg\mapsto\updatePcPerm{\reg(r)},
      \overline{r_{\mathit{arg}}} \mapsto \reg(\overline{r_{\mathit{arg}}}),r_0
      \mapsto c_{\mathit{ret}}, r_\stk \mapsto c_\stk, r \mapsto \reg(r)]$ 
    \item $\npair[n-1]{c_{\mathit{ret}}} \in \stdvr(W')$
    \item $\npair[n-1]{c_\stk} \in \stdvr(W')$
    \end{itemize}

    then we have that $\npair[n-1]{(\reg',\ms'')} \in \observations(W')$
  \item \label{item:scall:hyp-cont} \emph{Hyp-Cont} If
    \begin{itemize}
    \item $n' \leq n-2$
    \item $W'' \futurewk \revokeTemp{W}$
    \item $\memSat[n']{\ms''}{\revokeTemp{W''}}$ 
    \item for all $r$, we have that:
      \begin{equation*}
        \reg'(r)
        \begin{cases}
          = c_{\mathit{next}} &\text{ if } r = \pcreg\\
          = \reg(r)&\text{ if } r \in \overline{r_{\mathit{priv}}}\\
          \in \stdvr(\revokeTemp{W''}) &\text{ if $\reg'(r)$ is a global capability and } r \not\in \{\pcreg,\overline{r_{\mathit{priv}}}, r_\stk\}
        \end{cases}
      \end{equation*}
    \item $\reg' \text{ points to stack with $\ms_\stk$ used and $\ms_{\mathit{unused}}''$ unused}$ for some $\ms_{\mathit{unused}}''$
    \end{itemize}

    then we have that $\npair[n']{(\reg',\ms'' \uplus \ms_f \uplus \ms_\stk \uplus \ms_{\var{unused}}'')} \in \observations(W'')$
  \end{enumerate}
  Then 
  \begin{itemize}
    \item $\npair{(\reg,\ms \uplus \ms_f \uplus \ms_\stk \uplus \ms_{\var{unused}})} \in \observations(W)$
  \end{itemize}

  % Note: $k_{\mathit{act-rec}}$ is the (standard) size of an activation record (2
  % words for stack pointer and return pointer and some fixed number of words for
  % the standard return instructions).
\end{lemma}
The essence of the \texttt{scall} lemma is to show \texttt{scall} is executed in
a reasonable state (\ref{item:scall:memsat}-\ref{item:scall:stack}), it is safe
to actually transfer control to the callee (\ref{item:scall:hyp-callee}), and that
the return capability given to the callee is safe (\ref{item:scall:hyp-cont}),
which corresponds to showing that the remainder of the program is safe to execute.

In the common case \texttt{scall} is used to invoke untrusted code, and we will
only have very general assumptions about that code such as it has contains a
capability for a linking table with malloc in it and all other addresses are
instructions. These values are all safe which allows us to conclude that the
read condition holds for the untrusted code which allows us to use the
Theorem~\ref{thm:ftlr} to conclude that it can be safely executed if the memory
and the register-file are safe. By assumption in Hyp-Callee, the memory is safe,
so it suffices to show that the register-file contains safe values, which
amounts to showing that the arguments in the call are safe.

By using Lemma~\ref{lem:scall-works} to reason about \texttt{scall}, the proofs
become agnostic to the actual implementation of \texttt{scall}. In other words,
should we change the implementation of \texttt{scall}, then we just need to
prove that Lemma~\ref{lem:scall-works} holds for the new implementation in order
to gain that all our results still hold true.

The \texttt{malloc} instruction is also used often, so we also prove
Lemma~\ref{lem:malloc-works} to help reason about it. The structure of the
\texttt{malloc} lemma is close to that of the \texttt{scall} lemma. It also has
requirements on the configuration just before \texttt{malloc} is executed
(\ref{item:malloc:looking-at}-\ref{item:malloc:malloc-sat}) as well as
requirements on the execution afterwards (\ref{item:malloc:hyp-cont}). It does,
however, not have any requirements on the callee as we can rely on the
\texttt{malloc} specification to reason about this part once and for all.
\begin{lemma}[$\mathtt{malloc}$ works]
  \label{lem:malloc-works}
  If
  \begin{enumerate}
  \item \label{item:malloc:looking-at}\lookingat{(\reg,\ms)}{\mathtt{malloc}\;r\;k}{c_{\var{next}}}
  \item $k \geq 0$
  \item \linksto{(\reg,\ms)}{\malloc}{k}{c_\malloc}
  \item $c_\malloc$ satisfies the $\malloc$ specification with $\iota_{\malloc,0}$
  \item $W \futurestr [i \mapsto\iota_{\malloc,0}]$
  \item $\memSat{\ms}{W}$
  \item $\ms = \ms' \uplus \ms_{\var{footprint}}$
  \item \label{item:malloc:malloc-sat} $\memSat{\ms_{\var{footprint}}}{[i \mapsto W(i)]}$
  \item \label{item:malloc:hyp-cont} \emph{Hyp-Cont}
    If
    \begin{itemize}
    \item $n' \leq n-1$
    \item $\iota_\malloc \futurewk W(i)$
    \item $\memSat[n']{\ms_{\var{footprint}}' \uplus \ms'}{W[i \mapsto \iota_\malloc]}$
    \item $\memSat[n']{\ms_{\var{footprint}}'}{[i \mapsto \iota_\malloc]}$
      \begin{equation*}
        \reg'(r') = 
        \begin{cases}
          c_{\var{next}} & r' = \pcreg \\
          ((\rwx,\glob),\start,\addrend,\addr) & r' = r \\
          \reg(r) & r' \not\in \RegName_t \union \{\pcreg, r, r_1\}
        \end{cases}
      \end{equation*}
    \item $\addrend - \start = k - 1$
    \item $\dom(\ms_{\var{alloc}}) = [\start,\addrend]$
    \item $\forall \addr \in [\start,\addrend]\ldotp \ms_{\var{alloc}}(\addr) = 0$
    \end{itemize}
    Then we have $\npair[n']{(\reg',\ms' \uplus \ms_{\var{footprint}}'  \uplus \ms_{\var{alloc}})} \in \observations(W[\iota_\malloc])$
  \end{itemize}
  Then
  \[
    \npair{(\reg,\ms)} \in \observations(W)
  \]
\end{lemma}
In the technical appendix~\citep{technical_appendix}, we also define a lemma for
the macro used to create closures, \texttt{crtcls}. We stress that these lemmas
are conveniences and are not strictly necessary to reason about programs for the
capability machine. In principle, one could reason solely in the semantic model
without the use of these lemmas, but that would lead to a lot of repetition.
\end{toplas}

@@

\subsection{Reasoning about programs definitions}
\begin{definition}
  We say that $(\reg,\ms) \text{ is looking at } [i_0,\cdots,i_n] \text{ followed by } c_{\mathit{next}}$ 
  iff
  \begin{itemize}
  \item $\reg(\pcreg) = ((p,g),b,e,a)$
  \item $p = \rwx$, $p = \exec$, or $p = \rwlx$
  \item $a+n\leq e$, $b\leq a\leq e$
  \item $\ms(a+0,\cdots,a+n) = [i_0,\cdots,i_n]$
  \item $c_{\mathit{next}} = ((p,g),b,e,a+n+1)$
  \end{itemize}
\end{definition}

\begin{definition}
  We say that ``\linksto{(\reg,\ms)}{\var{key}}{j}{c_\malloc}'' 
  iff
  \begin{itemize}
  \item $\reg(pc) = \stdcap$
  \item $\ms(\start) = ((\_,\_),\start_\link,\_,\_)$
  \item $\ms(\start_\link+j) = c$
  \end{itemize}
\end{definition}

\begin{definition}
  We say that $\reg \text{ points to stack with $\ms_\stk$ used and $\ms_{\mathit{unused}}$ unused}$
  iff
  \begin{itemize}
  \item $\reg(r_\stk) =((\rwlx,\local),b_\stk,e_\stk,a_\stk)$
  \item $\dom(\ms_{\mathit{unused}}) = [a_\stk+1,\cdots,e_\stk]$
  \item $\dom(\ms_\stk) = [b_\stk,\cdots,a_\stk]$ \lau{Maybe make it clear what happens when $\ms_\stk$ is empty}
  \item $b_\stk - 1\leq a_\stk$
  \end{itemize}
\end{definition}
